\mychapter{6}{Conclusion}
In order to achieve the goal of the Paris agreement of limiting the global temperature increase to "well below" 2 °C, and increasing amount of carbon removal will be required. The UNEP estimates the annual "emissions gap" to be around 20 GtCO$_2$ by the year 2100. It is therefore imperative to develop a portfolio of negative emissions technologies that can be used to close this gap and reach a net-zero greenhouse gas effect. This thesis find that that no single NET can deliver Paris‑consistent CDR at acceptable cost. In contrast to current VCM narratives, often focused on projects with single technologies, only a diversified portfolio with shifting weights over time is capable of doing so.\vspace{2mm}

Regarding RQ1, the models find that each of the four approaches has a potential in the range of multiple gigatonnes of CO$_2$ per year, which is limited by available land and biomass for the nature-based NETs. The costs range from an average of 10 USD t$^{-1}$ for A/R all the way up to 282 USD for DAC.\vspace{2mm}

In response to RQ2, the results show that A/R and BiCRS are used for the near-term targets, while DAC carries the long-term burden in the second half of this century. Enhanced weathering is theoretically promising but severely restricted by logistical and social constraints. The resulting dynamic portfolio shows a total cost of 177 trillion USD over 75 years for CDR totaling 879 GtCO$_2$ at an average price of 202 USD t$^{-1}$. The central insight from this dynamic portfolio approach is that timing matters as much as cost. By explicitly modeling future NET requirements and technology learning, the dynamic approach moves beyond the price-driven voluntary carbon markets that currently dominate. A static approach, focused only on minimizing present cost would put "all eggs in one basket" by betting entirely on afforestation. However, A/R saturates within the coming decades as less additional land is available, while the engineered NET have no cumulative limits. The time effect is especially relevant for DAC, since its cost trajectory depends on learning effects, which in turn are defined by the cumulative deployment. Delaying investments in DAC therefore not only push back its availability, but also keep it expensive. Ultimately, no single NET can deliver the required scale on its own: The optimal combination requires all four technologies to be deployed in the right amount at the right time to fully exploit their respective potential.\vspace{2mm}

Considering RQ3, all relevant NET face limitations not only by technical but also by social and political constraints. This is especially problematic for ERW, which requires extensive mining operations, but DAC and BECCS are also expected to face resistance regarding CO$_2$ transportation and storage. These factors are not just secondary considerations, but, at least in the case of ERW, binding constraints that limit deployment to significantly below the technical potential.\vspace{2mm}

\clearpage
Finally, answering RQ4, policymakers can use these dynamic portfolios to design strategies and subsidy programs that incentivize investments in high-durability, scalable technologies (DAC, ERW, partially BECCS) while managing the transition away from low-quality offsets and land-limited options like A/R. Investors can use them as a roadmap for phased CDR development aligned with long-term net-zero targets. Durable, novel NETs, especially DAC, can act as long-term infrastructure bets, and those who enter the market earlier can capture the benefits of learning effects and future scarcity, but are exposed to regulatory risk if climate targets are weakened. To create certainty for CDR investments, governance frameworks that define quality standards for CRCs should be established, and existing carbon pricing and cap-and-trade systems, such as the EU ETS, should allow the use of high-quality carbon removals to offset corporate carbon balances. To incentivize NET investments, a variety of subsidies or even public procurement and retirement of CRC should be considered where existing voluntary carbon markets fail to direct sufficient investments to future-proof technologies.\vspace{2mm}

To conclude, this thesis has shown that in order to meet the 2 °C goal, using the various types of currently available, typically cheap carbon offsets and removals of doubtful quality and permanence is not sufficient. A strategic, timed deployment of a diversified portfolio of high-quality, permanent NETs is necessary to ensure that sufficient carbon removals are available to close the remaining emissions gap in the long term while balancing cost, scalability, and societal acceptance. Future work should combine such portfolios with policy design and financing instruments in order to create concrete deployment pathways.