\mychapter{3}{Techno-Economic Modeling of NETs}
In order to create optimal portfolios of negative emission technologies, the four relevant approaches were systematically modeled from a technological and economic perspective. The goal of these techno-economic models was to calculate the country-level availability and cost of these NETs in order to construct a global marginal removal cost curve (MRCC). These MACCs served as the answer to the first research question ("What is the techno-economic potential and cost range of the four selected NETs under realistic assumptions?"), and, more importantly, provided the basis for the optimization models used to generate optimal portfolios of negative emission technologies developed in the next chapter.

\section{Methodology}
Generally, a bottom-up process was used to construct techno-economic models of the four relevant NETs. For each NET, the entire process from feedstock or energy input to the final output of the carbon credits was modeled. This included the capital investments required to construct exemplary projects or plants, as well as the fixed and variable costs of operating these plants (i.e. materials, energy and labour). The models also included relevant leakage factors from production, transport and storage. In generic terms, the LCOC is calculated as
\begin{equation}
\text{LCOC} = w + f + c = w + \frac{\sum{PV(k^\text{fixed}_t)}}{\sum{q_t}} + \frac{\sum{I_tx_t}}{\sum{q_t}}*\text{CRF}
\end{equation}
where $w$ is the variable cost per tonne of CDR, $\sum{q_t}$ is the sum of CO$_2$ removed, $PV(k^\text{fixed}_t)$ is the present value of the annual fixed costs for operation and maintenance, $\sum{I_tx_t}$ is the system price (i.e., the capital investment) required for the project, and CRF is the capital recovery factor, which, at an assumed discount rate of 7\% and project lifetime of 20 years, is 9.44\%. For the other LCOC-derived terms, such as the LCOCR OR LCOCR$^{+*}$, the costs of transportation and storage are added, co-revenues deducted, or the leakage factor applied to the sum of CDR.\vspace{2mm}

For the input data, country-level data on the availability and, if relevant, cost of the primary constraints and inputs, such as land and biomass, was gathered from the FAOSTAT database \parencite{fao_faostat_2025, fao_faostat_2025-1, fao_faostat_2026}. Additionally, country-level data on the cost of labor, materials and other relevant inputs was collected. If such data was not available, global averages were calculated and adjusted for each country using purchasing power parity (PPP) data of the World Bank \parencite{world_bank_icp_2021}. The collected data was then fed into the techo-economic models in order to build a global marginal removal cost curve for each NET.

\include*{afrf}
\include*{bicrs}
\include*{ewr}
\include*{dac}

\mychapter{4}{Development of Optimal Portfolios}
To construct optimal CDR portfolios, two types of optimization approaches were considered. First, a simple static portfolio was created, which is based on a one-off removal of a fixed quantity of carbon dioxide on a fixed timeline, such as the removal of 5 Gt of CO$_2$ over five years. In the second step, this simple static approach was evolved into a dynamic portfolio approach, which considered a dynamically increasing annual requirement of CDR in order to achieve certain targets on a certain timeline, such as the annual removal of 10 Gt of CO$_2$ by the year 2050.\vspace{2mm}

For both types, the techno-economic models described in Section 3 were implemented in the Python programming language, using the Pyomo package and the Gurobi solver. For the portfolio optimization, multi‑period linear programs (LPs with SOS2) were used, for which the objective was to minimize the total costs, constrained by the availability of each NET and the requirement of fulfilling the respective CDR target.\vspace{2mm}

The BECCS model was used to representat for both BiCRS approaches. This simplification is usefule and valid because both BiCRS approaches are constrained by, and therefore compete for, the same biomass feedstock base, and the marginal removal cost curves of both technologies exhibit a similar cost and availability range. As a result, explicitly including both models would have primarily affected the allocation within the BiCRS category, but not materially impacted the overall portfolio results. This modeling choice reflects a complexity-conscious implementation of the BiCRS pathways, but does not imply that BECCS and Biochar are identical on a technological level. The cost of ERW was modeled using a piecewise linear function derived from the MACC to allow the model to choose any operating level and compute the resulting total costs without introducing nonlinearities. Since the capital expenditures of ERW make up less than 2\% of its LCOC, using the MACC as a fixed, exogenous input, instead of separately modeling the capital investments, is a sufficiently realistic approximation. The same approach was used for BECCS in the static models. For the dynamic BECCS model, the fixed and variable costs and the capital investments were modeled separately, meaning that the optimizer chose not only the operating level for BECCS, but also the timeline for capacity construction, similar to the DAC model. Piecewise functions were also applied for the rock use of ERW and the learning curve of DAC. In Pyomo, they were implemented via SOS2 piecewise constraints, which approximate nonlinear cumulative cost curves. The CapEx for DAC are defined by a bilinear function of the amount built in a given year and the learning-curve based cost per unit. A piecewise-constant step function was used as an approximation to preserve computational tractability. The A/R model uses the assisted natural regeneration approach (Section 3.2.2), is linear and therefore did not require the use of a piecewise approximation. \vspace{2mm}

The following restrictions were considered in the optimization models: A/R was restricted by the growth of trees, the total available land area, and a limitation on the annual reforested area relative to all available land. In addition, an annual reversal rate was implemented to simulate fires and pests. The BECCS model was limited by the available amount of waste biomass, an initial availability of electric power capacity, and an annual maximum capacity increase of, while the technology itself was assumed to be available at a constant construction cost. ERW was restricted by the agricultural land area available for spreading, as well as a limitation of annual mining capacity relative to the global coal mining industry, which increased linearly over the operating years. DAC was only limited by an assumed initial capacity and a maximum annual capacity increase relative to the previous year's capacity.

\include*{stat_port}
\include*{dyn_port}

\mychapter{5}{Discussion}

\section{Static Portfolios}

The static portfolios should be interpreted as cost‑minimizing under a one‑off removal constraint and the assumptions of Section 3, rather than as universally optimal long‑term strategies. Still, a number of conclusions can be drawn from the portfolios shown in Figures 4.1 and 4.2. First of all, the specific target formulation has a significant impact on the results. With the annual constraint, the removals must be delivered immediately, which forces the use of more costly BECCS removals to cover the gap left by cheaper but slower A/R and ERW. Because A/R has annual planting limits and a growth lag, it is not able to fully satisfy short-term targets on its own, especially not in the case of annual constraints. Enhanced weathering faces similar constraints regarding its expansion and use of rock material, and therefore plays only a very minor role, while BECCS is deployable much quicker.\footnote{However, it should be noted that the deployability of BECCS is partially due to the model design, which accepts its use as a "gap filler" that is necessarily able to achieve the annual target from the start.} Additionally, the annual constraint creates an overshoot effect for the 50 year time horizon: From a cost perspective, it is cheaper to plant an excessive amount of trees early on than to rely on the more expensive alternative NET instead, even if that leads to over-achieving the target during the later years. Without the annual removal constraint, the optimizer can delay removals from the first year to later times. This temporal flexibility allows it to rely nearly entirely on the cheaper A/R approach, and makes engineered CDR entirely disappear for the longer time horizons. Lastly, it is worth noting that DAC sees no application in either of the scenarios due to its high costs compared to the other NETs, making it non-competitive in the static optimization model.\vspace{2mm}

The stark difference in costs between the long and short time-frames is similarily explained not only by the increased use of cheaper A/R in the longer time-frames, but also the fact that A/R itself becomes cheaper the longer the trees are allowed to grow. This is due to the fact that the cost of A/R is largely defined by its upfront cost for forest restoration, while carbon removal occurs over the entire growth cycle of 30 to 50 years. In the longer time-frames, each tree therefore removes more carbon, making A/R more efficient and decreasing its cost by an order of magnitude.\vspace{2mm}

From the perspective of VCM participants, the static model makes A/R the most attractive approach, allocating most removals to it as long as land is available. For example, if a company wanted to achieve the goal of "historic carbon neutrality" by removing the sum of its past emissions in a one-off project, the model suggests that simply buying carbon removal credits from afforestation is the most economical solution.\vspace{2mm}

These results illustrate the limitations of a simple, static cost-based approach to selecting negative-emission technologies discussed in chapter 2 of this thesis. Under these static constraints, low-cost removals dominate the solution, especially if temporal flexibility is available, while novel NETs are used only to fill near-term gaps. The BECCS and ERW approaches serve as temporary "compliance technologies" and Direct Air Capture is not deployed at all. This result carries two issues common for such a static, present-cost based perspective: First, the potential cost reduction due to increased deployment, especially the learning effects of DAC, are entirely disregarded. However, even if cost reductions were considered, A/R would likely remain the cheapest option. The second, and more fundamental, issue is that this type of model treats the low-cost attractiveness of A/R as if it was indefinitely available, ignoring the limited nature of A/R, which is in reality constrained by the global land area available for forest restoration. Once this limited potential is exhausted, reliance on A/R alone is no longer possible. The static approach therefore underestimates the value of early investments in engineered NETs, which may be valuable not because they are the cost-optimal option at that time, but because they help drive down future costs and ensure long-term CDR availability once cheaper approaches hit their natural limits.

\section{Dynamic Portfolio}

In contrast to the static portfolios, the dynamic portfolio considers the inter-temporal trade‑offs between present cost, future availability of A/R and BECCS, and learning‑based cost reductions for DAC. The dynamic portfolio shows staggering numbers, both, in regard to the total amount of carbon removed until the year 2100, and in regard to the total investments and operating expenditures necessary to achieve this target. For comparison, the 879 Gt of CDR required over the next 75 years are approximately equivalent to the entire historic emission of humanity until 1990, and slightly less than the sum of emissions since 1990, which amounts to about 1000 GtCO$_2$ \parencite{ritchie_co_2020}.\vspace{2mm}

The development of the costs of the various technologies also provides interesting insights. The nearly flat cost curve of ERW is caused largely by the modeling constraints, which limit the actual deployment of the technology to a point significantly below the hypothetical potential shown by the MACC developed in Section 3.3. The increasing cost of BECCS is explained by the fact that the optimizer uses the cheapest potential first, e.g. the cheaply available manure. Once these options are are exhausted, it moves to add the less economical input materials. The initial cost decrease of A/R is caused by the delayed "delivery" of removals: The afforestation itself creates high upfront costs, which are recouped over time when carbon is absorbed into biomass. The later increase in A/R cost can be traced to the exhaustion of available land, forcing the optimizer to resort to more expensive options. Afforestation reaches its maximum potential in the late 2030s, while BECCS reaches its peak around 2050. While they will keep contributing significant amounts of CDR, their expansion will be stopped by natural limitations in land and waste biomass availability. Finally, the results show that A/R removals start to decline from the mid-2060s, when most available land is used up and maturing forests absorb less CO$_2$.\vspace{2mm}

Looking at DAC, the fact that its cost seems to only decrease slightly after the sharp initial drop can be traced to the nature of the learning effects: The amount of DAC capacity constructed under the early, high-cost conditions makes up only a small share of the total DAC capacity. By the time the 100th Mt of capacity is deployed, the cost will already be below 1000 USD, and by the first Gt, it will have fallen to about 600 USD. With continuing deployment, the cost decrease per unit will be less proncounced.\vspace{2mm}

\section{Interpreting the Techno-Economic Potential and Costs}
Returning to research question one, Table 5.1 provides an overview of the CDR potential and cost ranges of the four approaches modeled in this chapter. The models find that afforestation is the cheapest NET overall, with most of the potential being available at below 10 USD tCO$_2^{-1}$. However, this potential is limited by the land availability and biomass growth cycle. It therefore dominates in the short-term, but eventually saturates and needs to be replaced by more engineered removal approaches starting from the late 2060s.\vspace{2mm}

Enhanced weathering also shows enormous potential of nearly 9 GtCO$_2$ yr$^{-1}$ at costs below 70 USD tCO$_2^{-1}$, which is significantly cheaper than the remaining novel options. However, it is severely limited by the logistical constraints of building what is effectively a large-scale, global mining operation similar in size to the present-day coal mining industry. Therefore, despite its potential, ERW only makes up about 5\% of the cumulative removals under realistic assumptions, despite being cost-competitive.\vspace{2mm}

\begin{table}[h!]
\renewcommand{\arraystretch}{1.3}
\adjustbox{max width=\textwidth}{%
\begin{tabular}{lccc}
\hline
\multirow{2}{*}{\textbf{Technology}} 
& \multicolumn{2}{c}{\textbf{Cost (USD tCO$_2^{-1}$)}} 
& \multirow{2}{*}{\textbf{Potential (GtCO$_2$/yr)}} \\
\cline{2-3}
 & \textbf{Lower} & \textbf{Upper} & \\ \hline
\textbf{A/R}   & 3   & 88  & 2{,}431 \\
\textbf{BECCS} & 131 & 333 & 4{,}793 \\
\textbf{ERW}   & 26  & 67  & 8{,}630 \\
\textbf{DAC}   & --\footnotemark & 507 & -- \\ \hline
\end{tabular}}
\caption{Modeled cost ranges and potential of selected NETs}
\end{table}
\footnotetext{The cost of DAC declines with increasing deployment, asymptotically approaching its levelized operating cost.}

The BiCRS approach, represented here by the BECCS model, is critical in filling the intermediate period. With a potential of nearly 5 GtCO$_2$ yr$^{-1}$, it acts as a "bridge" that buys time for DAC to scale up sufficiently. Being available from 133 USD, it sits between A/R and ERW on the cheap end and DAC on the expensive side. Also, compared to other approaches, it is the only one with significant co-revenues, in this case from electricity generation, which significantly improve its economics.\vspace{2mm}

Finally, DAC is the most expensive of all options, but plays a critical role in the second half of the century, since it is the only option that has a technically unlimited potential for CDR, being limited virtually only by the available energy.\vspace{2mm}

Compared to the previous literature, as reviewed in Section 2, the cost and CDR potentials modeled generally fall within or close to the same ranges (see Table 2.2), with the exception of ERW, which was found to be available at a minimum cost about 50\% below the previous literature estimates. This implies that cost rankings are relatively robust, but absolute costs, especially for DAC, BECCS and ERW, are highly sensitive to assumptions about learning, logistics and MRV.

\section{The Optimal Cost-Minimizing Portfolio}
Research question two represents the core contribution of this thesis: The question of how the ideal, cost-minimizing portfolio of negative emissions technologies should be designed in order to meet the climate goals of the Paris agreement. The dynamic portfolio modeled in this thesis and presented in Figure 4.3 proposes one potential solution to this question. The proposed portfolio is unsurprisingly one that includes several technologies working together: Afforestation begins immediately due to its simplicity and low cost, while BECCS covers the requirements that A/R cannot meet due to natural limitations. Finally, DAC comes into operation in the mid-century and takes over the bulk of the CDR in the later decades. Compared to the static portfolio, it combines A/R, BECCS, ERW and DAC in a time‑staggered fashion rather than selecting a single "winning" technology.\vspace{2mm}

The actual critical findings, however, are the development of DAC capacity compared to its operations, and the role of ERW. DAC construction begins earlier than its LCOC curve, which only starts from the point where DAC commences operations in the mid-2040s. Investing in such plants, only for them to sit idle for nearly their entire lifetime, seems counterintuitive. However, it makes sense considering that operating them is still more expensive than using the other NETs as long as these can fulfill the annual CDR target. The initial DAC investment therefore occurs not to actually contribute to the CDR targets of those years, but to build up capacity and to be ready when it is required, while "buying down" the average capital expenditures per tonne of capacity. Effectively, the early DAC investments act as insurance against the future unavailability of sufficient CDR options come the saturation of cheaper NET. Therefore, these results should be regarded as an encouragement to begin DAC deployment early in order to drive down its costs by the time it is required in significant amounts from the mid-2040s onwards.\vspace{2mm}

The role of enhanced weathering is probably the biggest point of debate in these results. As the MACC developed before (Figure 3.9) shows, ERW has the potential to achieve about 9 Gt of CDR per year, which would alone be sufficient to cover nearly half of the annual CDR required by the year 2100. Yet, its role in the final portfolio is diminished to a maximum of 1 Gt annually, based on limitations of rock availability and logistics. This result can be interpreted in two ways: Either, one could argue that enhanced weathering is not worth pursuing due to the inherent operational problems. At the same time, looking at the cost of below 70 USD tCO$_2^{-1}$, it could be argued that all available efforts should be made to make this technology more available.\vspace{2mm}

The latter argument raises the question of the cost of this ideal portfolio. The total cost of 177 trillion USD over the next 75 years is so large that it significantly exceeds the total annual gross domestic product of the entire global economy, which is expected to reach 123 trillion USD in 2026 \parencite{imf_gdp_2026}. In other words, this portfolio would require 1.92\% of the global GDP per year to be spent on carbon removal. While this number doesn't seem that large, it should be noted that this would require every country in the world to contribute this amount to a coordinated, global effort. Considering the current state of international relations, which are increasingly dominated by opposing "spheres of influence", the success of such a coordinated efforts seems doubtful. Regardless of the way of its implementation, the investments that will be required to achieve sufficient carbon removal will probably have a huge impact on the global economy due to the diversion of investments away from economic growth and towards negative emissions technologies, which do not directly contribute to short-term value creation.

\section{Impact of Social and Political Constraints}
Looking at the high cost of the proposed portfolio and the fact that these technologies do directly not create economic value in themselves, the question of political and social feasibility arises. Although it can (and probably should) be argued that limiting climate change is of direct benefit to society and humanity as a whole, the long-term perspective and intangibility of this benefit may put such efforts at odds with the relatively short election cycles, especially of western democracies, and the short-term incentives they create for individual politicians seeking re-election.\vspace{2mm}

Additionally, each technology is likely to face individual opposition. The case of enhanced weathering discussed above is the starkest example of social and political limitations: Despite huge technical potential and relatively low cost, the impact of this approach is expected to be low due to the assumed resistance to the development of large-scale global mining operations, which would exceed global coal mining if the full potential of ERW was to be achieved. This suggests that social and political obstacles are not secondary considerations, but binding constraints that significantly afect the optimal portfolio. Besides ERW, other approaches also face obstacles: The general acceptance of BECCS, for example, is presumably much higher than that of ERW, but carbon transport through pipelines and long-term carbon storage could face significant local opposition, which is currently not considered in the optimization model. This issue naturally also concerns direct air capture. Additionally, DAC requires significant amounts of energy, which is a limited resource with intense competition from applications such as AI and electric vehicles of various types. A greatly increased energy demand from DAC could increase energy shortages and drive up prices, which may lead to opposition against using large amounts of energy for a technology that has little to no immediate, short-term benefits. Even A/R may be subject to opposition from those benefiting from deforestation, such as large-scale farmers using former rain-forests as agricultural land.\vspace{2mm}

In sum, it should be noted that a purely technical approach to CDR vastly diverges from a socially and politically feasible one. The present model attempts to capture some of the most pressing social and political concerns in order to present a realistic picture of CDR development. If policymakers and investors ignore these constraints and plan around technical potentials alone, they risk systematic under‑investment in DAC and over‑reliance on ERW that will likely not materialize at the assumed scale. Deeper analysis of social and political factors is required to ensure that the solutions proposed from a long-term perspective don't get derailed by the reality of politics and public opinion.

\section{Implications for Investors and Policy Makers}
Finally, the question arises as to which conclusions the two primary addressees of this study, the investors who either build or invest in NETs, and the policymakers responsible for the political framework of the global response to climate change, should draw from the results presented above.\vspace{2mm}

For investors, the primary learning is that most investments in engineered NET are strategic investments in the long-term future of carbon removal, and insurance against policy tightening around low‑quality credits. This is especially true for DAC, which is not economically competitive at the moment, but will become highly relevant in the second half of the 21st century. Considering the learning effects discussed above, investing now is an important step to secure the viability of DAC in the future, and those who can make it available at scale from the 2040s onwards will likely profit from an ever-increasing demand. Still, investors must consider that this projection and the results of this thesis rely on an expanding demand for CDR, which depends on political enforcement of the targets of limiting climate change set out in the Paris agreement. However, most political frameworks, such as the EU ETS, still have to implement the use of CRCs. Additionally, if unexpected changes in the volatile political environment led to a global abandonment of these targets and other goals of limiting climate change, investments in negative emissions would lose significant value and would rely solely on voluntary purchases by climate-conscious buyers.\vspace{2mm}

For policymakers, this means that simple carbon pricing is unlikely to deliver the required incentives, and that the most important instrument regarding negative emissions are laws and regulations that phase out low-quality offsets while incentivizing the use of high-quality CRCs, in order to create demand for durable CDR. Creating long-term security regarding climate change targets and the role which negative emissions will play in the future requires holistic frameworks to govern carbon markets, including, but not limited to, the quality of carbon removals. The EU Carbon Removals and Carbon Farming Regulation is a first step in this direction, but still does not differentiate between present price considerations and future capacity requirements. Additionally, as discussed in Section 2.3, politicians must also consider if and how NETs should be subsidized to create sufficient incentives for CDR investments, while maintaining the ongoing emission abatement efforts and convincing the public of the need of such public investments.

\section{Limitations}
The dynamic optimization model created in this thesis relies on the goal of limiting climate change agreed upon in the Paris agreement of 2015, and the carbon removal targets implied by this agreement, as set out by the UN Environment Programme Emissions Gap Report \parencite{unep_emissions_2017}. Faster decarbonization may significantly reduce the need for CDR, while slower decarbonization would increase it. A sensitivity analysis using various CDR targets would be useful to quantify these effects.\vspace{2mm}

Further, the models assume energy, especially electricity, to be available in unlimited quantities at a global average price. In a highly electrified, AI-driven future, energy cost and scarcity could significantly change the economics of NETs, especially DAC. Modeling various scenarios of energy supplies, both regarding availability and carbon intensity, would be a valuable addition. The dynamic model also assumes a learning rate of 10\% for DAC. If the realized rate is lower, DAC costs could remain prohibitively expensive, while a higher learning rate could make DAC more economical. Again, a sensitivity analysis would be valuable.\vspace{2mm}

The afforestation model does not consider the impact of climate change on future A/R efforts, which could go both ways: Increased CO$_2$ levels could increase biomass growth, but also cause destruction through climate change-induced natural disasters. Weighting the impact of climate change on A/R would be valuable to further understand the role it can play in future CDR efforts. The A/R cost model also does not consider the opportunity cost of afforestation, that is, the foregone profits from using the respective lands for other purposes, such as agriculture. Adding these costs would probably increase the modeled cost of A/R, especially for tropical rainforests, which provide the majority of A/R potential, and were removed to create economically valuable agricultural land.\vspace{2mm}

With respect to BiCRS, the model assumes that BECCS and Biochar are sufficiently similar to allow the use of BECCS to represent BiCRS in general. While this is true from a global perspective, the individual conditions on a country level may prefer the one technology over the other. For example, developing countries with little infrastructure could prefer Biochar, while industrialized nations might instead use BECCS due to economies of scale. Allowing such choice might result in a combined BECCS/Biochar MACC with higher availability or lower costs, and warrants further research into which BiCRS approach is ideal based on local conditions. The model also does not consider the future development of waste biomass supplies, which could increase, e.g., due to population growth or from increased forestry waste available by increased afforestation. Detailed projections of future biomass supplies would provide insights into how much NET portfolios can rely on BiCRS-based approaches in the future.\vspace{2mm}

Regarding economic modeling, the financing conditions and cost of capital are assumed to be the same globally. In a realistic scenario, these conditions would vary greatly between countries and adjusting them for political risk and economic conditions would probably result in higher rates, and consequently capital recovery factors, especially in developing countries. Additionally, taxes and depreciation-based tax advantages are not considered in the models. Considering financing conditions and tax benefits for NETs would probably reduce the relative cost gap between developed and developing nations.\vspace{2mm}

Finally, the model also excludes technologies such as direct ocean capture, ocean alkalinization, and other experimental approaches. While this is sensible considering their early stage and the resulting difficulty of creating reliable models, these technologies could become important with further research. Also, the BiCRS model uses only waste biomass in order to avoid raising the issue of land competition created by using purpose-grown energy crops grown. It would be valuable to consider how much additional biomass could be made available through dedicated plantation, since this would reduce the reliance on DAC in the second half of the century.\vspace{2mm}

These limitations imply that the portfolios presented in this chapter should be read as scenario‑based results rather than absolute forecasts, and that their optimality is still conditional on the technologies selected and assumptions made.
